Robotic Bicycle Balance Assistance Reduces Probability of Falling When Subjected to Mechanical Perturbations at Low Speeds

### 2024-04-08 main.tex
Added more text to the results and added a first take on the discussion.

\usepackage{amsmath}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}
\usepackage{todonotes}

\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
Subjected to Mechanical Perturbations}

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    Measure & Variable & Units & Sensor \\
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
  \end{tabular}
\end{table}

@@

\subsection{Statistics}
%
Our primary hypothesis is that the balance assistance controller will reduce
the probability of falling when perturbed externally at the handlebar. We have
a single binary outcome variable \(f\) that is dependent on several possible
explanatory independent variables including the binary balance assistance on or
off.

@@

\(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
which follows a Bernoulli distriction given the probability \(p_{ij}\) that a
fall occurs. The log-odds of the probabiliy is then a linear function of our
independent variables.

@@

All of variables are scaled such they they have a mean of zero and a standard
deviation of one. We use cluster-mean centering with the clusters being an
individual subject.

@@

\todo[inline]{Marten writes "The skill variable is not included, because it was
not collected correctly for all participants." This needs to be addressed.}

@@

The coefficient estimates for a single-level logistic regression at a speed of 6
\unit{\kilo\meter\per\hour} and gain $k=-10$ are shown in table \ref{table:freq-coefs-6}. The
associated multiplicative change in odds are shown in table \ref{table:freq-change-in-odds-6}.

\begin{table}
  \centering
  \caption{Coefficient estimates at 6 \unit{\kilo\meter\per\hour} and gain
  $k=-10$ based on frequentist single-level logistic regression.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lllll}
    \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
    Intercept                                     & -0.29             & 0.17                    & 0.09             & \\
    Angular impulse                               & 1.69              & 0.27                    & 0.00             & \\
    Perturbation order                            & -0.77             & 0.22                    & 0.00             & \\
    Balance-assist on                             & -0.64             & 0.27                    & 0.02             & \\
    Roll angle                                    & -0.25             & 0.21                    & 0.24             & \\
    Steer angle                                   & -0.14             & 0.21                    & 0.51             & \\
    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             & \\
    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             & \\
    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             & \\
    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             & \\
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=-10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-change-in-odds-6}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
  \end{tabular}
\end{table}

The coefficients estimates of a single-level logistic regression are shown in table
\ref{table:freq-coefs}. The associated multiplicative changes in odds are shown in table
\ref{table:freq-change-in-odds}.

\begin{table}
  \centering
  \caption{Coefficient estimates at 10 \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
      regression.} \label{table:freq-coefs}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
      Intercept                                     & -0.24             & 0.16                    & 0.13             & \\
      Angular impulse                               & 2.39              & 0.29                    & 0.00             & \\
      Perturbation order                            & -1.16             & 0.21                    & 0.00             & \\
      Balance-assist on                             & -0.44             & 0.24                    & 0.07             & \\
      Roll angle                                    & 0.27              & 0.22                    & 0.22             & \\
      Steer angle                                   & -0.37             & 0.24                    & 0.12             & \\
      Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08             & \\
      Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11             & \\
      Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29             & \\
      Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25             & \\
  \end{tabular}
\end{table}

@@

>>>>

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}
\usepackage{todonotes}

\def\kph{\kilo\meter\per\hour}

\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
Subjected to Mechanical Perturbations at Low Speeds}

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary outcome variable \(f\) that is dependent on several possible
explanatory independent variables including the binary balance assistance (on
or off).

@@

\(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
which follows a Bernoulli distribution given the probability \(p_{ij}\) that a
fall occurs. The log-odds of the probabiliy is then a linear function of our
independent variables with \(\beta\) as the intercept and \(\alpha_k\) as the
linear coefficients.

@@

We scalle all independent variables such they they have a mean of zero and a
standard deviation of one. We use cluster-mean centering, as recommended by
\cite{Enders2007}, with the clusters being an individual subject because we are
interested in the association between the state of the balance-assist system
and the outcome of the perturbation.

Cluster-mean centering shows there to be no variation between participants,
i.e. that rider skill (the only participant differentiating factor) is not a
predictor. We expected there to be a variation between participants in how well
they are able to resist the perturbation. However, this was not true. This
allowed us to use a simpler single-level logistic regression model, instead of
a multilevel model. This left use with angular impulse, perturbation order,
balance assist state, and roll & steer angles at the time of perturbation. We
also include interaction effects between the balance assist state and the other
four variables.

\todo[inline]{Marten writes "The skill variable is not included, because it was
not collected correctly for all participants." This needs to be addressed. What
about skill being tied to participant clusters?}

@@

@@

The coefficient estimates for a single-level logistic regression at
6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in Table
\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and balance
assist state are all statistically significant predictors. Larger angular
impulse increases the probability to fall and enduring more perturbations or
having the balance assist on, decrease the probability to fall.  The associated
multiplicative change in odds are shown in
Table~\ref{table:freq-change-in-odds-6}.

\begin{table}
  \centering
  \caption{Logisitic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $k=10$.}
  \label{table:freq-coefs-6}
  \begin{tabular}{llll}
    \toprule
    Variable                                      & Estimate          & Standard error          & p-value          \\
    \midrule
    Intercept                                     & -0.29             & 0.17                    & 0.09             \\
    Angular impulse                               & 1.69              & 0.27                    & 0.00             \\
    Perturbation order                            & -0.77             & 0.22                    & 0.00             \\
    Balance assist state                          & -0.64             & 0.27                    & 0.02             \\
    Roll angle                                    & -0.25             & 0.21                    & 0.24             \\
    Steer angle                                   & -0.14             & 0.21                    & 0.51             \\
    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             \\
    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             \\
    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             \\
    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             \\
    \bottomrule
  \end{tabular}
\end{table}

\todo[inline]{Combine the columns of the multiplicative change in odds into the
coefficient table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-change-in-odds-6}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kilo\meter\per\hour} with gain $k=8$ are shown in Table
\ref{table:freq-coefs}. The angular impulse and perturbation order are
statistically significant predictors. Larger angular impulse increases the
probability to fall and enduring more perturbations decreases the probability
to fall. The associated multiplicative change in odds are shown in
Table~\ref{table:freq-change-in-odds}.

\begin{table}
  \centering
  \caption{Logisitic regression coefficient estimates at
    10~\unit{\kilo\meter\per\hour} and gain $k=8$.}
  \label{table:freq-coefs}
  \begin{tabular}{llll}
    \toprule
    Variable                                      & Estimate          & Standard error          & p-value \\
    \midrule
    Intercept                                     & -0.24             & 0.16                    & 0.13    \\
    Angular impulse                               & 2.39              & 0.29                    & 0.00    \\
    Perturbation order                            & -1.16             & 0.21                    & 0.00    \\
    Balance-assist on                             & -0.44             & 0.24                    & 0.07    \\
    Roll angle                                    & 0.27              & 0.22                    & 0.22    \\
    Steer angle                                   & -0.37             & 0.24                    & 0.12    \\
    Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08    \\
    Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11    \\
    Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29    \\
    Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25    \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Balance-assist}

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a fall occurs while cycling at a speed of 6~\si{\kph} but at
10~\si{\kph} that is not the case \((p=0.07)\). If all other factors remain the
same at 6~\si{\kph}, the balance-assist system halves the odds that a
perturbation results in a fall. How much this changes the probability that a
fall occurs depends on the values of the other predictor variables: if the odds
that a perturbation results in a fall are a 1000:1, turning on the
balance-assist system reduces the odds to 500:1. However, the probability that
a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability. For example, if the odds that a fall occurs
is two, halving it to one reduces the probability from $0.66$ to $0.50$. Figure
\ref{fig:probability} shows how the balance-assist system changes the fall
probability for different magnitudes of angular impulse.

\begin{figure}
  \centering
  \subcaptionbox{6~\si{\kph}}{
  \includegraphics[width=70mm]{example-image-a}
    %\includegraphics[width=\textwidth]{figures/predicted_fall_probability-6.png}
  }
  \subcaptionbox{10~\si{\kph}}{
    \includegraphics[width=70mm]{example-image-b}
  }
  \caption{Comparison of predicted fall probability for balance-assist on or
    off All other predictor variables are fixed at zero, except for the
    interaction effect between the balance-assist and the angular impulse.}
  \label{fig:probability}
\end{figure}

Angular impulse magnitude has the largest significant effect for predicting
fall probability. An increase in angular impulse increases the fall probability
both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
more important predictor at higher speeds compared to lower speeds. The reason
for this likely has to do with the width of the treadmill. As a bicycle travels
at higher speeds, the same perturbation causes larger lateral deviations. At
10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
of the treadmill. If the same experiment was performed on an infinite plane,
the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
could often recover in the alloted treadmill width. Our results are very much
dependent on the two modes of falling with use: exit the treadmill width or
foot is placed on the belt. Cycle paths are a similar width as the treadmill,
so ride's are often limited in width when recoverying from a fall.

\todo[inline]{Could show an impulse response in lateral deviation for different
speeds.}

An increase in the number of perturbation that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed.

Roll and steer angle are not a significant predictor of fall probability,
neither at 6 or at 10~\si{\kph}. We expected this to have an effect. If you are
in a rolled and steered state that is far from the upright equilibrium, then a
perturbation that further pushes you from the equilibrium should have some
additive or multiplicative effect on the resulting motion trajectory.

None of the interaction effects are statistically significant. That means that
whether the balance-assist system is on or off does not significantly change
the effect that the roll angle, steer angle, angular impulse, and perturbation
order have on the probability that a fall will occur.
